\chapter{Conclusion}

\textit{This dissertation has discussed the researching, implementation and evaluation of several semi-supervised models for use with genomic
data. It has also discussed techniques for processing data and improving model performance, and has used the lessons learned from 
this to construct a simple command line tool.}

\section{Results}

This project contains a comparison of semi-supervised models on genomic data, showing that semi-supervised learning can be used to 
improve phenotype classification accuracy. The other outputs of this project are: a publicly available repository of semi-supervised models in PyTorch; 
a methdology for evaluating these models; and a simple tool for users without machine learning experience to perform semi-supervised 
learning.

\section{Success Criteria}

All the major success criteria of the project have been achieved. 

The implementations of the models performed as well as their original paper implementations on the MNIST dataset, showing that they were 
implemented correctly. All the models are now available online at \url{https://github.com/Clondon98/Semi-Supervised_Models}, and include the 
best performing PyTorch implementation of the semi-supervised VAE and ladder that I was able to find online. This repository also allows 
the models to be easily used on different datasets, while other implementations are restricted to MNIST. 

The M2 and ladder models also succeeded in achieving statistically significant performance improvements over the fully supervised model on 
gene expression data, and an ensemble of both of them achieved even higher performance. This led to the development of a 
command line tool that can be trained and used for classification very simply, with no need for knowledge of machine learning techniques.

I also managed to implement an extension that calculated the saliency of inputs to the network, and while I ran out of time to include it 
in the tool, the results on MNIST show that saliency maps can be used to calculate and visualise feature importance.

\section{Further Work}

Some areas of the project that could be explored further include:
\begin{itemize}
    \item Using VAEs to generate additional unlabelled datapoints and adding these to the dataset to see if they improve performance.
    \item Using a semi-supervised VAE to generate additional labelled datapoints and seeing if these help improve supervised learning 
          performance.
    \item Discovering the most important genes for phenotypes using saliency and comparing those to highly important genes found through
          other methods.  
\end{itemize}

\section{Final remarks}

I hope that the models used in this project can be refined further to find real use in the fields of medicine and biology. The models and tool
developed in this project are all freely available online and will hopefully be of use to future researchers interested in semi-supervised
learning.