\chapter{Introduction}

\textit{The stated aim of this project was to develop a semi-supervised autoencoder-based method for classifying 
cells into phenotypes \footnote{the observable characteristics and traits of an organism} using genetic data. To this end a range 
of autoencoder based semi-supervised models have been implemented, 
ranging from fairly simple to state-of-the-art. These models were then evaluated on selected datasets, including 
the Cancer Genome Atlas expression data.}

\section{Motivation}

Classifying organisms into phenotypes using genomic data is important in both medicine and agriculture. In medicine uses include 
classifying tissues as cancerous and non-cancerous~\cite{Li2017}, predicting whether a strain of \textit{Helicobacter pylori} will lead to 
stomach cancer~\cite{helicobacter}, and diagnosing people more susceptible to Crohn's disease and bipolar disorder~\cite{doi:10.1002/humu.23280}.
In agriculture it can be used to predict whether a crop will grow well under harsh conditions before the crop is planted~\cite{cimmyt}

In this project I will attempt to classify into phenotypes based on gene expression data. Gene expression is the process of synthesising 
proteins from the gene via RNA. A transcriptome is the set of all RNA 
molecules in a cell or population of cells. While every cell with a nucleus in an organism has the same DNA and genes, the cells
differ in function, and this depends on which genes are being actively expressed. This in turn means the
transcriptome contains information from both genetic and epigenetic sources~\cite{Gibney2010}. Epigenetic differences are differences
in the phenotype without alterations to the DNA (e.g. DNA methylation), and therefore gene expression data provides more information about 
the phenotype than DNA seqeuncing.

These classifications and predictions are typically computed using a supervised learning technique, which requires that all the 
datapoints used in training be associated with a phenotype label. However, biological labs worldwide perform transcriptome analysis, 
generating gene expression data 
which is available online. This, along with the development
of RNA-Seq using next generation sequencing has resulted in huge amounts of gene expression data being available for researchers. However,
while many of the datasets generated in different experiments may include transcriptomes for the desired organism,
the majority of the time the experiments are measuring different phenotypes of the organism. This means that a 
researcher wanting to analyse or predict a specific phenotype is unable to use much of the available data, being limited
to only those labelled with the desired phenotype.

Semi-supervised learning attempts to leverage unlabelled data to improve the accuracy of the machine learning
algorithm on a supervised task. Autoencoders have a long history of being used in semi-supervised learning problems,
having been used early on to improve deep networks by pretraining them using stacked denoising autoencoders (Section~\ref{sdae})
and recently having been used to achieve state of the art semi-supervised performance as part of the ladder 
network (Section~\ref{ladder}). The main reason for their use is that autoencoders are good at learning 
important features of data in an unsupervised manner, and these features are often useful in improving 
supervised performance. Therefore, with the use of semi-supervised autoencoder models it should be possible to leverage the data without
the desired phenotype to improve performance in predicting the phenotype.

The reason for choosing autoencoder-based models is that they are implemented
using neural networks. This is advantageous because neural networks work well on non-linear data and are
also able to effectively analyse data with high dimensionality (number of features). Transcriptomes can
often contain several thousand genes, and so any model used must be able to cope with this level of dimensionality.

This project is also the first comparison between these semi-supervised models on gene expression data, and is one of very few
studies to have performed semi-supervised deep learning on genomic data.

\section{Related work}

Stacked denoising autoencoders have previously been used with gene expression data to derive the most informative
genes for distinguishing between healthy and cancerous cells~\cite{8217828}.

Likewise, variational autoencoders (Section~\ref{vae}) have been successfully used to extract a biologically relevant latent 
space from cancer transcriptomes ~\cite{Way2018ExtractingAB}. They and the semi-supervised variant (Section~\ref{ssVAE}) 
have also been used to model the change in the gene expression of tumours in response to certain drugs~\cite{10.1093/bioinformatics/btz158}.
he semi-supervised model in the paper, Dr.VAE, jointly models both the drug response and the treatment outcomes.

Ladder networks have also been used in biologically relevant (non semi-supervised)
ways, achieving state of the art accuracy in the binary cancer classification
problem using gene expression data~\cite{10.1007/978-3-319-78723-7_23}.